Divergent morphology of elliptical bifaces from sites in the American Southeast articulates with raw material differences

### 2022-08-05 supp/index.Rmd
<edit index>

Elliptical bifaces from the Millsap Cache and Jowell Farm (41AN13) sites comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area. Similar to Gahagan bifaces, elliptical bifaces from Jowell Farm are assumed to have been manufactured from Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade with central Texas groups. Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material. The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---biface shape in the small class, and to test the hypothesis that greater morphological variation would be apparent in this class due to the variable idiosyncracies related to resharpening and/or retouch. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding suggests a shared mental template, coupled with a size differential driven by raw material differences. Additionally, size does not differ between raw materials in the large class; however, in the small class, it does. This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through their uselife. In following with expectations, greater shape diversity was found to occur in the small class, where bifaces from Jowell Farm are more diverse than those from the Millsap Cache. Discrete raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.

>>>>

Elliptical bifaces from the Millsap Cache and Jowell Farm (41AN13) sites comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area. Similar to Gahagan bifaces, elliptical bifaces from Jowell Farm are assumed to have been manufactured from Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade with central Texas groups. Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material. The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---biface shape in the small class, and to test the hypothesis that greater morphological variation would be apparent in this class due to idiosyncratic responses related to resharpening and/or retouch. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding suggests a shared mental template, coupled with a size differential driven by raw material differences. Additionally, size does not differ between raw materials in the large class; however, in the small class, it does. This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter use at a similar shape and size, but diverge significantly in size---but not shape---through that uselife. In following with expectations, greater shape diversity was found to occur in the small class, where bifaces from Jowell Farm are more diverse than those from the Millsap Cache. Discrete raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.
